\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2021-07-28 paper.tex

The association structure between the longitudinal and the time-to-event submodels is key to the joint modeling framework. In spite of that, rationale for selecting shared associations has received little attention. We then propose to include some of the most common parameterization with no \textit{a priori}, set out in Table~\ref{table:shared_associations}, and let the model select the relevant ones through the regularization strategy described in Section~\ref{sec:penalized-obj}.
\begin{table}[htb]
\centering
\begin{tabular}{ccccc}
\toprule
Description & $\varphi_a(t, \beta_k^l, b^l)$ & $\imath_a$ & Reference \\
\midrule
Linear predictor & $m_k^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
Random effects & $b^l$ & $r_l$ & \citet{hatfield2011joint} \\
Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
 \bottomrule
\end{tabular}
\caption[]{Description of the shared associations included in the group-specific risk of event submodel~\eqref{eq:intensity-model}.}
\label{table:shared_associations}
\end{table}

@@

Now, for the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
\begin{equation*}
  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]
\end{equation*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{i, t_j, \gamma_k^{(w)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
The update for $\beta_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-beta_k}
  \beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k) + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}
The previous trick we used to handle the non-differentiability of the $\ell_1$ part in the elastic net is no longer sufficient for~\eqref{eq:minimization-beta_k}, since the $\ell_2$ norm in the sparse group lasso is not squared, leading to a non-differentiability at 0 (which is precisely the criterion exploited by the penalty to set some groups of coefficients to exactly 0~\citep{simon2013sparse}). Then, we choose to solve problem~\eqref{eq:minimization-beta_k} using the proximal gradient descent~\citep{boyd2004convex} described in Algorithm~\ref{alg:ISTA}, where we first define
\[p_{\varrho}(y) = \argmin_u \Big\{\dfrac{1}{2\varrho} \left\lVert u - \big(y - \varrho  \nabla f(y) \big)\right\rVert_2^2 + \zeta \norm{u}_{\text{sg} l_1, \tilde{\eta}} \Big\}.\]

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\beta_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:beta-update}
  \beta_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}(\beta_k^{(w)}, R^{(w)}_{n,k}, \tilde \eta, \zeta_{2,k}).
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}.
The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
\begin{align}
  \label{eq:grad-beta_k}
  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
\end{align}
for all $l = 1, \ldots, L$. Note that the gradients of the considered $\varphi_a$ functions with respect to $\beta_k^l$ are given in Table~\ref{table:shared_associations}.

@@

\paragraph*{The QNMCEM algorithm.}
Algorithm~\ref{alg:QNMCEM} describes the main steps of the resulting QNMCEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}. The $Condition$ in algorithm~\ref{alg:QNMCEM} is cv$(\Delta_{\text{rel}}^{(w+1)}) >$ cv$(\Delta_{\text{rel}}^{(w)})$, where cv$(\Delta_{\text{rel}}^{(w)})$ is coefficient of variation at the $(w + 1)$-th iteration computed as
\begin{equation*}
  \label{eq:coefficient_variation}
  \text{cv}(\Delta_{\text{rel}}^{(w+1)})= \dfrac{\text{sd}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})}{\text{mean}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})},
\end{equation*}
where sd(.) and mean(.) are the sample standard deviation and mean functions respectively and relative difference $\Delta_{\text{rel}}^{(w+1)}$ is given by 
\begin{equation*}
  \label{eq:relative_difference}
  \Delta_{\text{rel}}^{(w+1)} = \text{max} \Big\{\dfrac{|\theta^{(w+1)} - \theta^{(w)}|}{|\theta^{(w)}| + \epsilon } \Big\}
\end{equation*}

\begin{algorithm}[]
\DontPrintSemicolon
\caption{QNMCEM algorithm for the lights model inference}
\label{alg:QNMCEM}
\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}    
{ 
  \tcc{Monte Carlo E-step}
  Compute $\hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g_i$\\
  Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
  \If{Condition}
  {
      $N \leftarrow N + $\\
      $tol \leftarrow tol - $
  }
  \tcc{Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:beta-update} \\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}

>>>>

The association structure between the longitudinal and the time-to-event submodels is key to the joint modeling framework. In spite of that, rationale for selecting shared associations has received little attention. We then propose to include some of the most common parameterization with no \textit{a priori}, set out in Table~\ref{table:shared_associations}, and let the model select the relevant ones through the regularization strategy described in Section~\ref{sec:penalized-obj}.
\begin{table}[htb]
\centering
\begin{tabular}{ccccc}
\toprule
Description & $\varphi_a(t, b^l)$ & $\imath_a$ & Reference \\
\midrule
Linear predictor & $m_k^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
Random effects & $b^l$ & $r_l$ & \citet{hatfield2011joint} \\
Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
 \bottomrule
\end{tabular}
\caption[]{Description of the shared associations included in the group-specific risk of event submodel~\eqref{eq:intensity-model}.}
\label{table:shared_associations}
\end{table}

\subsection{Graphical model for the lights}
Figure~\ref{fig:graphical-model} below gives the graphical representation for the lights model to get a beter view of the dependence structure for the model.
\begin{figure}[!htb]
\centering
\resizebox{.9\textwidth}{!}{
\begin{tikzpicture}
\hspace{3cm}
\tikzstyle{main}=[circle, minimum size = 18mm, ultra thick, draw =black!80, node distance = 13mm,text centered]
\tikzstyle{var}=[rectangle, minimum size = 16mm, ultra thick, draw =black!80, node distance = 13mm,text centered]
\tikzstyle{connect}=[-latex, thick]
\tikzstyle{box}=[rectangle, draw=black!100]
  \node[var, fill = white!100] (beta) [] {\Large$\bm{\beta_k^l}$};
  \node[var, fill = white!100] (xi) [below=4cm of beta] {\Large$\bm{\xi_k}$};
  \node[var, fill = white!100] (gamma) [below=3cm of xi] {\Large$\bm{\gamma_k}$};
  \node[main, fill = white!100] (G) [right=4cm of xi] {\Large$\bm{G_i}$};
  \node[main, fill = white!100] (mu) [right=4cm of beta] {\Large$\bm{M_{ik}^l(t)}$};
  \node[main, fill = black!30] (Y) [right=of mu] {\Large$\bm{Y_i^l(t)}$};
  \node[main, fill = white!100] (b) [below=0.7cm of Y] {\Large$\bm{b_i^l}$};
  \node[var, fill = white!100] (phi) [above=of Y] {\Large$\bm{\phi_l}$};
  \node[var, fill = white!100] (Sigma) [right=of b] {\Large$\bm{D}$};
  \node[main, fill = white!100] (W) [below=of b] {\Large$\bm{\varphi_i(t)}$};
  \node[main, fill = white!100] (lambda) [right=4cm of gamma] {\Large$\bm{\lambda_i(t)}$};
  \node[main, fill = black!30] (T) [right=of lambda] {\Large$\bm{T_i}$};
  \node[main, fill = black!30] (X) at (3, -8.3) {\Large$\bm{X_i}$};

  \path [->,draw,ultra thick]
        (beta) edge (mu)
        (G) edge (mu)
        (b) edge (mu)
        (mu) edge (Y)
        (Sigma) edge (b)
        (phi) edge (Y)
        (xi) edge (G)
        (X) edge (G)
        (gamma) edge (lambda)
        (W) edge (lambda)
        (G) edge (lambda)
        (X) edge (lambda)
        (lambda) edge (T)
        ;

  \node[rectangle, inner sep=8mm,draw=black!100, ultra thick, fit= (beta) (xi) (gamma),label={[anchor=west, yshift=-0.5cm, xshift=-10mm]{\Large$\bm{k = 1...K}$}}] at (0, -4.6) {};
  \node[rectangle, inner sep=4mm,draw=black!100, ultra thick, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi)] {};
  \node[rectangle, inner sep=4mm, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi),ultra thick] at (12, 3.5) {\Large$\bm{l = 1...L}$};
  \node[rectangle, dashed, inner sep=3mm, draw=red!100, fit= (b) (W),ultra thick] {};

  \node[rectangle,inner sep=6mm, draw=black!100, label=center:\rotatebox{90}{\color{white}\Large{\textbf{Longitudinal}}}, minimum width=1cm, minimum height = 6cm, fill = black!50] at (-2.7,0.5) {};

  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:\rotatebox{90}{\color{white}\Large{\textbf{Time-to-event}}}, minimum width=1cm, minimum height = 6cm, fill = black!50] at (-2.7,-8.2) {};

\end{tikzpicture}}
\caption{Graphical representation of the lights model. Square boxes denote observed data while circles denote unobserved (including random) terms. The red-dashed box indicates that the association features and the random effects are correlated, which, in addition to the link with the latent group membership, gives rise to the joint model.}
\label{fig:graphical-model}
\end{figure}

@@

Now, for the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
\begin{equation*}
  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]
\end{equation*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
The update for $\beta_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-beta_k}
  \beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k) + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}
The previous trick we used to handle the non-differentiability of the $\ell_1$ part in the elastic net is no longer sufficient for~\eqref{eq:minimization-beta_k}, since the $\ell_2$ norm in the sparse group lasso is not squared, leading to a non-differentiability at 0 (which is precisely the criterion exploited by the penalty to set some groups of coefficients to exactly 0~\citep{simon2013sparse}). Then, we choose to solve problem~\eqref{eq:minimization-beta_k} using the proximal gradient descent~\citep{boyd2004convex} described in Algorithm~\ref{alg:ISTA}, where we first define
\[p_{\varrho}(y) = \argmin_u \Big\{\dfrac{1}{2\varrho} \left\lVert u - \big(y - \varrho  \nabla f(y) \big)\right\rVert_2^2 + \zeta \norm{u}_{\text{sg} l_1, \tilde{\eta}} \Big\}.\]

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\beta_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:beta-update}
  \beta_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}(\beta_k^{(w)}, R^{(w)}_{n,k}, \tilde \eta, \zeta_{2,k}).
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}.
The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
\begin{equation}
  \label{eq:grad-beta_k}
  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
\end{equation}
for all $l = 1, \ldots, L$.

@@

\paragraph*{The QNMCEM algorithm.}
Algorithm~\ref{alg:QNMCEM} describes the main steps of the resulting QNMCEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.

% \vspace{.5cm}

\begin{algorithm}[!htb]
\DontPrintSemicolon
\caption{QNMCEM algorithm for the lights model inference}
\label{alg:QNMCEM}
\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}    
{ 
  \tcc{Monte Carlo E-step}
  Compute $\hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g_i$\\
  Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
  \If{Condition}
  {
      $N \leftarrow N + $\\
      $tol \leftarrow tol - $
  }
  \tcc{Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:beta-update} \\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}
Let us define the relative difference at step $(w+1)$ as
\[\Delta_{\text{rel}}^{(w+1)} = \max_{j=1,\ldots,\vartheta} \Big\{\dfrac{|\theta^{(w+1)}_j - \theta^{(w)}_j|}{|\theta^{(w)}_j| + \epsilon_1 } \Big\}.\]
Then in practice, the convergence for Algorithm~\ref{alg:QNMCEM} is declared when the condition
\[\Delta_{\text{rel}}^{(w+1)} < \epsilon_2\]
is satisfied for 3 consecutive iterations, since the monotonicity property is lost due to the Monte Carlo approximations.
